---
title: "hw02-tianqi-lu"
author: "Tianqi Lu"
date: "9/30/2017"
output: github_document
---

# Import the data with base
```{r}
data <- read.csv("/Users/TLuv/Downloads/stat133/stat133-hws-fall17/nba2017-player-statistics.csv", header = TRUE, colClasses = c("character", "character", "factor", "character", "double", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer", "integer"))
```
```{r}
str(data)
```
```{r}
library("readr")
data <- read_csv("/Users/TLuv/Downloads/stat133/stat133-hws-fall17/nba2017-player-statistics.csv", col_types = list(col_character(), col_character(), col_factor(NULL), col_character(), col_double(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer(), col_integer()))
```
```{r}
str(data)
```

# 3. Right After importing the data
```{r}
data$Experience[data$Experience == "R"] = 0
data$Experience <- as.integer(data$Experience)
```

# 4. Performance of Players
```{r}
Missed_FG <- data$FGA - data$FGM
Missed_FG
Missed_FT <- data$FTA - data$FTM
Missed_FT
PTS <- 2 * data$Points2 + 3 * data$Points3 + data$FTM
REB <- data$DREB + data$OREB
MPG <- data$MIN / data$GP
EFF <-  (PTS + REB + data$AST + data$STL + data$BLK - Missed_FG - Missed_FT - data$TO) / data$GP
summary(EFF)
```

## Add the columns to the data frame
```{r}
EFF <- round(EFF, 5)
data <- cbind(data, Missed_FG, Missed_FT, PTS, REB, MPG, EFF)
```

```{r}
hist(EFF, main = "Histogram of Efficiency (EFF)", xlab = "EFF", ylab = "Frequency", col = "grey")
```

## Display the player name, team, salary, and EFF value of the top-10 players by EFF in decreasing order
```{r}
a <- round(head(sort(data$EFF, decreasing = TRUE), n = 10),5)
data[match(a, data$EFF), c("Player", "Team", "Salary", "EFF")]
```

## Provide the names of the players that have a negative EFF
```{r}
data[data$EFF < 0 ,"Player"]
```
## Use the function cor() to compute the correlation coefficients between EFF and all the variables used in the EFF formula
```{r}
corPTS <- cor(data$EFF, data$PTS)
corREB <- cor(data$EFF, data$REB)
corAST <- cor(data$EFF, data$AST)
corSTL <- cor(data$EFF, data$STL)
corBLK <- cor(data$EFF, data$BLK)
corMFG <- -cor(data$EFF, data$Missed_FG)
corMFG
corMFT <- -cor(data$EFF, data$Missed_FT)
corMFT
corTO <- -cor(data$EFF, data$TO)
corGP <- cor(data$EFF, data$GP)
```
## Display the computed correlations in descending order, either in a vector or a data frame. And create a barchart with the correlations (bars in decreasing order) like the one below.
```{r}
a <- data.frame(c(corPTS, corREB, corAST, corSTL, corBLK, corMFG, corMFT, corTO))
b <- cbind(a, color = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE))
colnames(b) = c("corr", "color")
barplot(b$corr, main = "Correlations between Player Stats and EFF", names.arg = c("PTS", "REB", "STL", "AST", "BLK", "Missed_FT", "Missed_FG", "TO"), cex.names = 0.7, col = c("red", "grey")[(b$color > 0) + 1])
```
# 5. Efficiency and Salary
Once you’ve calculated the Efficiency statistic, produce a scatterplot between Efficiency
(x-axis) and Salary (y-axis), including a lowess smooth line (locally weighted scatterplot
smoothing). Also, compute the linear correlation coefficient between them. What can you
say about the relationship between these two variables?
```{r}
plot(x = data$EFF, y = data$Salary, xlab = "Players' efficiency", ylab = "Salary")
lines(lowess(data$EFF,data$Salary), col = "blue")
cor(data$EFF, data$Salary)
```
According to the graph, we can draw a conclusion: Generally, the more efficient a player is, the more he gets paid. But there are many players having high efficiency but getting paid little.

- Taking into account the column MPG (minutes per game) select those players that have
an MPG value of 20 or more minutes per game
```{r}
players2 <- data[data$MPG >= 20, ]
```
- Create a data frame players2 with these players
```{r}
players2 <- data.frame(players2)

```
- Use this data frame to create a scatterplot between Efficiency and Salary, including a
lowess smooth line.
```{r}
plot(players2$EFF, players2$Salary, xlab = "Players with 20 or more MPG efficiency", ylab = "Salary")
lines(lowess(players2$EFF, players2$Salary), col = "red")
```
- Compute the linear correlation coefficient between these variables.
```{r}
cor(players2$EFF, players2$Salary)
```
- What can you say about the relationship between these two variables for the set of
“more established players”?

It's more proper to say the more efficient a player is, the more he gets paid.