---
title: "Lab 4: Reading Data Tables"
subtitle: "Stat 133, Fall 2017"
author: "Tianqi Lu"
output: html_document
---

> ### Learning Objectives
>
> - Importing Data Tables in R
> - Default reading-table functions
> - R package `"readr"`
> - R package `"foreign"`

------

## Importing Data Tables

Data sets come in many different presentations. One common format is that of a 
data table---like a spreadsheet---although you can find data in other formats 
(especially when the data is in its raw version). However, once you get to the 
analysis stage, you will likely be dealing with some sort of tabular format.

Because data tables are so ubiquituous, it is fundamental that you learn how to
import them in R. In the first part of today's lab, we are going to review 
various aspects that have to do with reading in tables in R.

We will talk about the following functions (and packages)

- `read.table()`
- `read.csv()`
- `read.delim()`
- `read.fwf()`
- R package `"readr"`
- `readHTMLTable()` from the R package `"XML"`
- Packages to read data sets from other software programs

-----

## Abalone Data Set

The first data set to consider is the __Abalone Data Set__ that is part of the
[UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Abalone)

The location of the data file is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data)

The location of the data dictionary (description of the data) is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names)

Look at both the dataset file, and the file with its description, and answer 
the following questions:

- What's the character delimiter?

Comma.

- Is there a row for column names?

No. But they are specified on another page.

- Are there any missing values? If so, how are they codified?

There's no missing value.

- What is the data type of each column?

Numeric.


One basic way to read this file in R is by passing the url location of the 
file directly to any of the `read.table()` functions:

```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"
abalone <- read.table(url, sep = ",")
```

### Getting a Local Copy of the Data

My suggestion when reading datasets from the Web, is to always try to get a local
copy of the data file in your machine (as long as you have enough free space
to save it in your computer). To do this, you can use the function `download.file()`
and specify the url address, and the name of the file that will be created in
your computer. For instance, to save the abalone data file in __your working 
directory__, type the following commands:

```{r}
# download copy
origin <- 'http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data'
destination <- 'abalone.data'
download.file(origin, destination)
```

Now that you have a local copy of the dataset, you can read it with 
`read.table()` like so:

```{r}
# reading data from your working directory
abalone <- read.table("abalone.data", sep = ",")
```

Once you read a data table, you may want to start looking at its contents,
usually taking a peek at a few rows. This can be done with `head()` and/or 
with `tail()`:

```{r}
# take a peek of first rows
head(abalone)

# take a peek of last rows
tail(abalone)
```

Likewsie, you may also want to examine how R has decided to take care of the
storage details (what data type is used for each column?). Use the function
`str()` to check the structure of the data frame:

```{r}
# check data frame's structure
str(abalone, vec.len = 1)
```


### Detailed information about the columns 

So far we have been able to read the data file in R. But we are missing a few 
things. First, we don't have names for the columns. Second, it would be nice
if we could specify the data types of each column instead of letting R guess
how to handle each data type.

According to the description of the Abalone data set, the columns represent 
these variables:

| Name           | Data Type  |
|----------------|------------|
| Sex		         | nominal    |
| Length		     | continuous |
| Diameter	     | continuous |
| Height		     | continuous |
| Whole weight	 | continuous |
| Shucked weight | continuous |
| Viscera weight | continuous |
| Shell weight   | continuous |
| Rings		       | integer    |

Let's create a vector of columns names, and another vector of data types:

```{r}
col_names <- c(
	'sex',
	'length',
	'diameter',
	'height',
	'whole_weight',
	'shucked_weight',
	'viscera_weight',
	'shell_weight',
	'rings'
)

col_types <- c(
	'factor',
	'numeric',
	'numeric',
	'numeric',
	'numeric',
	'numeric',
	'numeric',
	'numeric',
	'integer'	
)
```

Now we can re-read the table in a more complete (and usually more efficient) way:

```{r}
abalone <- read.table(
	'abalone.data',
	col.names = col_names,
	colClasses = col_types,
	sep = ","
)

# check its structure again
str(abalone, vec.len = 1)
```


### Your turn

- Read the Abalone data with the `read.csv()` function
- Look at the data description

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names) and confirm the following statistics:

```
       Length Diam  Height  Whole  Shucked  Viscera    Shell    Rings
Min	   0.075  0.055	0.000   0.002	 0.001    0.001	   0.002	    1
Max	   0.815  0.650	1.130   2.826	 1.488    0.760	   1.005	   29
Mean   0.524  0.408	0.140   0.829	 0.359    0.181	   0.239	9.934
SD	   0.120  0.099	0.042   0.490	 0.222    0.110	   0.139	3.224
```

```{r}
abalone_csv = read.csv("abalone.data", sep = ",", col.names = col_names, colClasses = col_types)
abalone_stat = data.frame(row.names = c("Min", "Max", "Mean", "SD"),
                          Length = c(min(abalone$length), max(abalone$length), mean(abalone$length), var(abalone$length)),
                          Diam = c(min(abalone$diameter), max(abalone$diameter), mean(abalone$diameter), var(abalone$diameter)),
                          Height = c(min(abalone$height), max(abalone$height), mean(abalone$height), var(abalone$height)),
                          Whole = c(min(abalone$whole_weight), max(abalone$whole_weight),
                                    mean(abalone$whole_weight), var(abalone$whole_weight)),
                          Shucked = c(min(abalone$shucked_weight), max(abalone$shucked_weight),
                                      mean(abalone$shucked_weight), var(abalone$shucked_weight)),
                          Viscera = c(min(abalone$viscera_weight), max(abalone$viscera_weight),
                                      mean(abalone$viscera_weight), var(abalone$viscera_weight)),
                          Shell = c(min(abalone$shell_weight), max(abalone$shell_weight),
                                    mean(abalone$shell_weight), var(abalone$shell_weight)),
                          Rings = c(min(abalone$rings), max(abalone$rings), mean(abalone$rings), var(abalone$rings)))
abalone_stat
```


### Basic Plots

As you can tell, the Abalone data contains 9 variables.
To start exploring the content, we begin by producing charts for each
single variable, focused on looking at their distributions:

- Quantitative variables: histogram, boxplot
- Qualitative variables: barchart, piechart

The workhorse plotting function in R is `plot()`. This function is actually
a _method_, meaning that it behaves differently depending on the type of input.

If the provided input is a factor, `plot()` will generate a barchart:

```{r}
# plot of a factor
plot(abalone$sex)
```

Alternatively, you can always create a frequency table first---via `table()`---
and then plot a barchart with `barplot()`:

```{r}
table_sex <- table(abalone$sex)
barplot(table_sex)
```

For a quantitative variable, the typical graphics to examine the distribution
are histograms (`hist()`) and boxplots (`boxplot()`):

```{r}
hist(abalone$diameter)
boxplot(abalone$diameter, horizontal = TRUE)
```

-----

## Pittsburgh Bridges Data Set 

The data set is part of the UCI Machine Learning Repository:

[http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges](http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges)

The data Description is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names)

The Data file is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1)


Read the description, and take a look at the data set:

- Are there column names?
- What is the field separator?
- Are there any missing values?
- What is the character for missing values (if any)?
- What is the data type of each variable (i.e. column)?
- Download a copy of the data to your computer (use `download.file()`)
and save it in a file named `bridges.data`

```{r}
# download a copy of the data file
download.file("http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1", "bridges.data.version1" )
```


### Reading the Data

- Create a vector of column names
- Create a vector of column types
- Use the function `read.table()` to read the data. Name it `bridges`.

```{r}
# vector of column names
column_names <-  c("IDENTIF", "RIVER", "LOCATION", "ERECTED", "PURPOSE", "LENGTH", "LANES", "CLEAR-G", "T OR D", "MATERIAL", "SPAN", "REL-L", "TYPE")

# vector of column types
column_types <-  c("character","factor","factor", "integer", "factor", "character","factor","factor","factor","factor","factor","factor","factor")

# reading the data with 'read.table()'
bridges <- read.table(
	'bridges.data.version1',
	col.names = column_names,
	colClasses = column_types,
	sep = ","
)

# reading the data with 'read.csv()'
bridges_csv <- read.csv(
	'bridges.data.version1',
	col.names = column_names,
	colClasses = column_types,
	sep = ","
)
```

### Using `read.csv()`

Now use the function `read.csv()` to re-read the bridges data
```{r}
# your code
bridges_csv <- read.csv(
	'bridges.data.version1',
	col.names = column_names,
	colClasses = column_types,
	sep = ","
)
```



### Basic Inspection

Use functions to start examining the `bridges` data frame:

- `str()`
- `summary()`
- `head()` and `tail()`
- `dim()`
- `names()`
- `colnames()`
- `nrow()`
- `ncol()`

```{r}
# your code
str(bridges_csv, vec.len = 1)
summary(bridges_csv)
head(bridges_csv)
tail(bridges_csv)
dim(bridges_csv)
names(bridges_csv)
colnames(bridges_csv)
nrow(bridges_csv)
ncol(bridges_csv)
```


### Want to do more?

Write R code to find:

- Year of the oldest bridge
- Year of the most recent erected bridge
- Frequency of bridges by purpose
- Frequency of materials
- Average length of the bridges
- Plot a timeline: year -vs- length

```{r}
# oldest bridge
bridges_csv[which.min(bridges_csv$ERECTED), 1]
# most recent
bridges_csv[which.max(bridges_csv$ERECTED), 1]
# frenquency of purpose
table(bridges_csv$PURPOSE)
# frenquency of material
table(bridges_csv$MATERIAL)
# average length
mean(as.integer(replace(bridges_csv$LENGTH, bridges_csv$LENGTH == "?", 0)))
# plot Year vs Length
plot(bridges_csv$ERECTED, bridges_csv$LENGTH, xlab = "Year", ylab = "Lenth", las = 1, col = "pink")
```

-----

