---
title: "hw03-tianqi-lu"
author: "Tianqi Lu"
date: "10/15/2017"
output: github_document
---

# Rankng of Teams

The analysis stage of this assignment has to do with looking at various ways to rank the
teams. Use an Rmd file for this part of your project.


## Basic Rankings

Start by ranking the teams according to salary, arranged in decreasing order. Use ggplot() to create a barchart (horizontally oriented), like the one shown below. The vertical red line is the average team salary.


```{r}
# Load needed libraries and create the table
library(scales)
library(ggplot2)
library(readr)
library(dplyr)
teams <- read_csv('../data/nba2017-teams.csv')

# Plot the chart
ggplot(teams, aes(x = reorder(team, salary), y=salary)) +
  geom_bar(stat='identity', fill = 'grey') +
  geom_abline(intercept = mean(teams$salary), slope = 0, 
              color = alpha('coral1', 0.7), size = 2)+
  coord_flip()+
  labs(y = 'Salary (in millions)', x = 'Team', 
       title = 'NBA Teams ranked by Total Salary')+
  theme_minimal()
```

The entire ranking is 
```{r}
select(arrange(teams, desc(salary)), team)
```


Create another bar chart of teams ranked by total points. The vertical red line is the average team points.


```{r}
library(scales)
ggplot(teams, aes(x = reorder(team, points), y=points)) +
  geom_bar(stat='identity', fill = 'grey') +
  geom_abline(intercept = mean(teams$points), slope = 0, 
              color = alpha('coral1', 0.7), size = 2)+
  coord_flip()+
  labs(y = 'Total Points', x = 'Team', 
       title = 'NBA Teams ranked by Total Points')+
  theme_minimal()
```


The entire ranking is 
```{r}
select(arrange(teams, desc(points)), team)
```


Use efficiency to obtain a third kind of ranking, and create an associated bar chart of teams ranked by total efficiency. The vertical red line is the average team efficiency.


```{r}
library(scales)
ggplot(teams, aes(x = reorder(team, efficiency), y = efficiency)) +
  geom_bar(stat='identity', fill = 'grey') +
  geom_abline(intercept = mean(teams$efficiency), slope = 0, 
              color = alpha('coral1', 0.7), size = 2)+
  coord_flip()+
  labs(y = 'Total Efficiency', x = 'Team', 
       title = 'NBA Teams ranked by Total Efficiency')+
  theme_minimal()
```


The entire ranking is 
```{r}
select(arrange(teams, desc(efficiency)), team)
```



## Principal Components Analysis (PCA)


Perform a principal components analysis (PCA) on the following variables, to use the first principal component (PC1) as another index to rank the teams:


```{r}
# select the variables that we are interested in from teams
focus <- teams[, -c(1, 2, 3, 7, 15)]
# create pca object (containnig all the PC's)
pca <- prcomp(focus, scale. = TRUE)
# add PC1 PC2 back to teams
teams <- mutate(teams, pc1 = pca$x[ ,1], pc2 = pca$x[ ,2])
# rearrrange team by PC1 in descending order
select(arrange(teams, desc(pc1)), team)
```


Createa a data frame with the eigenvalues:

```{r}
eigs <- data.frame(
  eigenvalue = round(pca$sdev^2, 4),
  prop = round(pca$sdev^2 / sum(pca$sdev^2), 4),
  cumprop = round(cumsum(pca$sdev^2 / sum(pca$sdev^2)), 4)
  )
eigs
```



```{r}
ggplot(data = teams, aes(pc1, pc2, label = team)) + 
  geom_point() + 
  geom_text(check_overlap = TRUE)
```
## Principal Components Analysis (PCA)


Perform a principal components analysis (PCA) on the following variables, to use the first principal component (PC1) as another index to rank the teams:


```{r}
# select the variables that we are interested in from teams
focus <- teams[, -c(1, 2, 3, 7, 15)]
# create pca object (containnig all the PC's)
pca <- prcomp(focus, scale. = TRUE)
# add PC1 PC2 back to teams
teams_with_pc <- mutate(teams, pc1 = pca$x[ ,1], pc2 = pca$x[ ,2])
# rearrrange team by PC1 in descending order
select(arrange(teams_with_pc, desc(pc1)), team)
```


Createa a data frame with the eigenvalues:

```{r}
eigs <- data.frame(
  eigenvalue = round(pca$sdev^2, 4),
  prop = round(pca$sdev^2 / sum(pca$sdev^2), 4),
  cumprop = round(cumsum(pca$sdev^2 / sum(pca$sdev^2)), 4)
  )
eigs
```


Use the first two PCs to get a scatterplot of the teams

```{r}
ggplot(data = teams_with_pc, aes(pc1, pc2, label = team)) + 
  geom_text(check_overlap = TRUE) +
  ggtitle('PCA plot (PC1 and PC2)')
```


In order to build an index based on the first PC, you are going to transform PC1. To get a more meaningful scale, you can rescale the first PC with a new scale ranging from 0 to 100. Once you have obtained the rescaled PC1, you can produce a barchart like the previous ones.


```{r}
# rescale PC1
teams_with_pc = mutate(teams_with_pc, 
               rs_pc1 = 100 * (teams_with_pc$pc1 - min(teams_with_pc$pc1)) 
               / (max(teams_with_pc$pc1) - min(teams_with_pc$pc1)))

# create a barchart from the rescaled pc1
ggplot(teams_with_pc, aes(x = reorder(team, rs_pc1), y = rs_pc1)) +
  geom_bar(stat='identity', fill = 'grey') +
  coord_flip()+
  labs(y = 'First PC (Scaled from 0 to 100', x = 'Team', 
       title = 'NBA Teams ranked by Scaled PC1') +
  theme_minimal()
```

Use the scaled PC1 to rank the teams:

```{r}
select(arrange(teams_with_pc, desc(rs_pc1)), team)
```






