---
title: "hw01-tianqi-lu"
author: "Tianqi Lu"
date: "9/20/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("scatterplot3d")
library(scatterplot3d)
```
# 1. A bit of data processing
## load data file
```{r}
load("data/nba2017-salary-points.RData")
ls()
```
## make a new salary variable s.t. the unit is million
```{r}
nsalary <- round(salary / 1000000, digits = 2)
```

## make a new experience variable s.t. 'R' is replaced with '0', and make it an integer vector
```{r}
indices <- experience == "R"
nexperience <- replace(experience, indices, "0")
nexperience <- as.integer(nexperience)
```

## new position
```{r}
nposition <- factor(position)
levels(nposition)[levels(nposition) == "C"] <- "center"
levels(nposition)[levels(nposition) == "PF"] <- "power_fwd"
levels(nposition)[levels(nposition) == "SF"] <- "small_fwd"
levels(nposition)[levels(nposition) == "PG"] <- "point_guard"
levels(nposition)[levels(nposition) == "SG"] <- "shoot_guard"
```

## compute frequency in position factor
```{r}
table(nposition)
```

# 2. Scatterplot of salary and points
```{r}
plot(points, nsalary, xlab = "Points", ylab = "Salary (in millions)", col = adjustcolor("chocolate1", alpha=0.5), cex = 1, pch = 19, main = "Scatterplot of Points and Salary")
```

# 3. Correlation between Points and Salary
## Mean of points
```{r}
xsum <- 0
n <- length(points)
xbar <- sum(points) / n
xbar
```
## Mean of salary(in millions)
```{r}
ysum <- 0
ybar <- sum(nsalary) / n
ybar
```

## variance of points(x) and SD(x)
```{r}
xvarsum <- 0
varX <- sum((points - xbar)^2)/(n-1)
varX
sdX <- sqrt(varX)
sdX
```

## variance of salary(in millions)(y) and sdY
```{r}
yvarsum <- 0
varY <- sum((nsalary - ybar)^2)/(n-1)
varY
sdY <- sqrt(varY)
sdY

```
## calculate covariance(x,y)
```{r}
covsum <- 0
covxy <- sum((points - xbar)*(nsalary - ybar))/(n-1)
covxy
```

## correlation(x,y)
```{r}
corxy <- covxy / (sdX * sdY)
corxy
```

# 4. Simple linear regression
```{r}
b1 <- corxy * sdY / sdX
b0 <- ybar - b1 * xbar
yhat <- b0 + b1 * points
yhat
```

## Questions
```{r}
summary(yhat)
```


- Regression equation is $\hat {y} = b_0 + b_1 * points$
- Slope coefficient $b_1$ stands for how much more salary when "points" increase by 1.
- Intercept term $b_0$ stands for the salary for 0-point players
- Predicted salary(in millions) for a player that scores:
  1. 0 points: `r b0`
  2. 100 points: `r b0 + b1 * 100`
  3. 500 points: `r b0 + b1 * 500`
  4. 1000 points: `r b0 + b1 * 1000`
  5. 2000 points: `r b0 + b1 * 2000`
  
  
# 5. Plotting the regression line
## 
```{r}
plot(nsalary~points,pch = 19, col = adjustcolor("chocolate1", alpha=0.5), main = "Scatterplot of Points and Salary", xlab = "Years of experience", ylab = "salary (in millions)")
abline(b0, b1, lwd = 2, col = "red" ) 
lines(lowess(points,nsalary), col = "blue")
text(x = c(2400, 2400), y = c(20, 30), labels = c("regression", "lowess"))
```

# 6. Regression residuals and coefficient of determination $R ^ 2$

```{r}
ei <- nsalary - yhat
RSS <- sum(ei^2)
TSS <- sum((nsalary - ybar)^2)
R2 <- 1-RSS/TSS
```

# 7. Exploring position and experience
```{r}
plot(nsalary~nexperience,pch = 19, col = adjustcolor("black", alpha=0.5), main = "Scatterplot with lowess smooth"); lines(lowess(nexperience,nsalary))
pllot <- scatterplot3d(x = points, y = nexperience, z = nsalary, pch = 16, color = adjustcolor("chocolate1", alpha=0.5) , main = "3d_relationship", ylab = "experience", zlab = "salary")

```
## Boxplot
```{r}
boxplot(nsalary ~ nposition, ylab = "Salary (in million)", xlab = "Position")

```
## Questions
- Provide concise descriptions for the plots of this section.
    This section gives us the relation between players' salary, players' positions and their experience. Boxplot gives the relation between position and salary. Scatterplot gives the relation between experience and salary, and 3d scatterplot gives the relation among these three.
- From the scatterplots, does Experience seem to be related with Salary?
    Yes. Generally the more experience you have, the more salary you get.
- From the boxplot, does Position seem to be related with Salary?
    Not really. Because the means are similar, the 25 quantiles are similar and the 75 quantiles don't vary a lot either.
    
# 8. Comments

The hardest parts in my mind are boxplot and lowess line. I know how to code it but don't fully understand what they stand for and how we use them.